---
title: "Lab 12 - Statistics, Coordinates, Facets, and Themes"
author: "George Rhodes"
date: "November 21, 2017"
output:
  pdf_document: default
  html_document: default
---

Complete the following exercises below. Knit together the PDF document and commit both the Lab 12 RMDfile and the PDF document to Git. Push the changes to GitHub so both documents are visible in your public GitHub repository.
```{r, echo = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(scales)
library(ggthemes)


#load National Survey on Drug Abuse and Mental Health and create object nsduh2015.
load("/Users/george/Documents/School/UW/SOC321/honors_thesis/honors_thesis/NSDUH-2015-survey-data.rda")

#change all variable names to all lowercase for ease of calling.
names(PUF2015_102016) <- tolower(names(PUF2015_102016))

#subset to relevant variables. 
subset_nsduh2015 <- PUF2015_102016 %>%
  select(questid2, filedate, txevrrcvd, alclottm, catag6, irsex, newrace2, irmaritstat, eduhighcat,
         al30est, alcbng30d, irpinc3, irfamin3, poverty3, coutyp2) 
#rename columns
colnames(subset_nsduh2015) <- c("ident", "date", "treatment", "alcohol", "age", "sex", "race", "marital", "edu", 
                                "drink30", "binge30", "income", "famincome", "poverty", "countytype")

#adds column 16 (treated_sober) with recovered, treated_drinking, and untreated 
data_treated_sober <- subset_nsduh2015 %>%
  mutate(treated_sober = ifelse((treatment == 1 & alcohol == 93), "recovered",
                                 ifelse(treatment ==1 & alcohol != 93, "treated_drinking",
                                     "untreated")))

#adding columns that count
# Treated and Sober "recovered"
data_count_recovered_n <- data_treated_sober %>%
  mutate(recovered_n = sum(treated_sober == "recovered"))

#Treated and Sober "treated_drinking""
data_count_not_recovered_n <- data_count_recovered_n %>%
  mutate(treated_drinking_n = sum(treated_sober == "treated_drinking"))

count_untreated_n <- data_count_not_recovered_n %>%
  mutate(untreated_n= sum(treated_sober == "untreated"))

data_clean <- count_untreated_n

#creates subet of only those "recovered"
recovered_respondants <- data_clean %>%
  filter(treatment == 1 & alcohol == 93)

#creates subset of only those "treated drinking"
treated_drinking_respondants <- data_clean %>%
  filter(treated_sober == "treated_drinking")

untreated_respondants <- data_clean %>%
  filter(treated_sober == "untreated")
                             

#following attempt to rename the values of data_clean$income failed to produce histogram
#x needs to be continuous, not discrete
data_clean$income[data_clean$income == 1] <- "0-9,999"
data_clean$income[data_clean$income == 2] <- "10,000-19,999"
data_clean$income[data_clean$income == 3] <- "20,000-29,999"
data_clean$income[data_clean$income == 4] <- "30,000-39,999"
data_clean$income[data_clean$income == 5] <- "40,000-49,999"
data_clean$income[data_clean$income == 6] <- "50,000-74,999"
data_clean$income[data_clean$income == 7] <- "75,000 +"

# More attempts to make this work.
# Also to make discrete so that density line doesn't dip down.
# oldvalues <- c("1", "2", "3", "4", "5", "6", "7")
# newvalues <- factor(c("0-10", "10-20", "20-30", "30-40", "40-50", "50-75", "75"))
# data_clean$income <- factor(data_clean$income)

#selects ident, treatment, alcohol, edu, income, famincome, poverty, and county type. 
project_variables <- recovered_respondants %>%
  dplyr::select(ident, treatment, alcohol, edu, income, famincome, poverty, countytype)
```

1. Choose one or more graphics you created for Lab 11 and either experiment with the underlying statistical layer if it already has one (i.e. if you made a histogram experiement with different bin widths) or add a separate statistical layer to your plot (i.e. a smooothing curve). Choose something you think will offer meaningful insight and describe why you made the choice you did. What additional information does this provide viewers of your graphic?

```{r, echo = FALSE, message = FALSE}
hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_bar(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_bar(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_bar(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket")

#Great. Just need to dodge. 
# can merge data set. use .id to call individual. .id saves which table data came from. 
  #one of last merges we covered in lecture. Use BIND. Row Bind. BOOM. bind_rows. 
# .id = name of table. Then specify color. then position = "dodge"
                  
                   

hist_income

# make color an attribute. Take out of ggplot. add fill = red for each geom_bar. 
hist_famincome <- ggplot(data_clean, 
                      mapping = aes(x = famincome, color = treated_sober)) +
                   geom_bar(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_bar(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_bar(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5)+
                   scale_y_continuous(labels = percent) +
                   labs(title = "Family Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket")
                    geom_density()

hist_famincome
```

I have spent so much time cleaning, reworking, and failing to seperate each bar within the histogram. I don't have time to finish this lab. The above visualizations are the best I have so far. I still need to seperate each bar, and find a way to add the income brackets. Both of which are proving very difficult. 

I am unable to change the binwidth and don't really understand where it goes. 
I will want to change the width of the bars once they are seperated to make this more aesthetically pleasing. I wouldn't mind a smoove curve connecting them, but have failed many times while playing with various stat functions. 

2. With the same or a different plot created in Lab 11, experiment with zooming in on specific areas of your graphic and changing the aspect ratio. Are their any benefits/drawbacks with either or both of these approaches for the visualizations you've created? What are they? 

```{r, echo = FALSE, message = FALSE}
hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   coord_cartesian(ylim = c(0, 25))
              
hist_income
```

```{r, echo = FALSE, message = FALSE}
hist_famincome <- ggplot(data_clean, 
                      mapping = aes(x = famincome, color = treated_sober)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5)+
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5)+
                   scale_y_continuous(labels = percent) +
                   labs(title = "Family Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket")+
                   coord_cartesian(ylim = c(0, 0.20))
hist_famincome
```

Once my visualizations show 3 distinct bars for each income bracket, I don't think I will not need to zoom in/out or change the aspect ratio. However, this is helpful and painless.


3. Try facetting a plot you have made by another categorical variable in your data (this can even be as simple as Male/Female). What is the difference between `facet_wrap()` and `facet_grid()`? How might facetting be useful in data visualization?

```{r, echo = FALSE, message = FALSE}
hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)
hist_income
```

4. Use the `theme()` layer to change the appearance of a plot of your choice including the
  * plot, axes, and legend titles
  * axes tick marks
  * text size
  * legend position
  
```{r, echo = FALSE, message = FALSE}
hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)+
                   theme_minimal()
hist_income

```
  

5. Create three versions of a graphic of your choice using different built-in themes or a theme created from `ggthemes`. Which ones do you think are best for presenting in an academic journal? A poster session? What are the qualities of the themes that you choice that you think make them more appropriate for presentation?

```{r, echo = FALSE, message = FALSE}

hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)+
                   theme_minimal()
hist_income

hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)+
                   theme_dark()
hist_income

hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)+
                   theme_pander()
hist_income

hist_income <- ggplot(data_clean, 
                      mapping = aes(x = income, color = treated_sober), stat = "count", breaks = seq(0, 7, by = 1)) +
                   geom_histogram(aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = recovered_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge", alpha = 0.5) +
                   geom_histogram(data = treated_drinking_respondants, 
                                  aes(y = ((..count..)/sum(..count..))), position = "dodge",alpha = 0.5) +
                   scale_y_continuous(labels = percent) +
                   labs(title = "Individual Income Distribution", x = "Annual income($)", y = "Percent of population in income bracket") +
                   facet_grid(.~treated_sober)+
                   theme_solarized()
hist_income
```


